### 2022-06-22 manuscript.qmd
init code for simulating DPM

---
title: "Causal inference for trait co-evolution using dynamic phylogenetic models"
format: html
editor: visual
---

## Introduction

Co-evolution can be identified statistically as non-zero between-trait phylogenetic correlations----or any other measure of mutual information over evolutionary time. Trait co-evolution is of interest in ecology and evolution when phylogenetic correlations are thought to reflect directed, causal relationships between traits, or when correlated traits share common causes.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-06-22 chop_tree.R

@@ -1,0 +1,38 @@
+require(phytools)
+require(phangorn)
+
+chop_tree <- function(tree) {
+  #### Cut up phylogenetic tree into segments #################
+  times <- node.depth.edgelength(tree) # date of each node
+  
+  # Line up date of each node with the split points in the tree
+  split_points <- sort(unique(times))
+  node_time <- match(times, split_points)
+  
+  # Create a sequence of nodes, respecting temporal order
+  node_seq <- seq(from = 1, to = length(node_time))
+  node_seq <- node_seq[order(node_time)]
+  
+  # Find the "parent" node for each node and tip of the tree
+  parent <- Ancestors(tree, node_seq, type = "parent")
+  
+  # Parent time indicates amount of time since the parent node, scaled by the total depth of the tree
+  parent_time <- rep(NA, length(node_seq))
+  #parent_time[1] <- -99 # placeholder for ancestral state
+  
+  for (i in 2:length(parent_time)) {
+    parent_time[i] <- (node.depth.edgelength(tree)[node_seq[i]] - node.depth.edgelength(tree)[parent[i]]) / max(node.depth.edgelength(tree))
+  }
+  
+  N_seg <- length(node_seq) # total num segments in the tree
+  
+  return(
+    list(
+      N_seg = N_seg,
+      node_seq = node_seq,
+      parent = parent,
+      ts = parent_time
+      )
+  )
+}
+

### 2022-06-23 chop_tree.R
add scaling argument to chop_tree

@@ -1,7 +1,7 @@
 require(phytools)
 require(phangorn)
 
-chop_tree <- function(tree) {
+chop_tree <- function(tree, scale=T) {
   #### Cut up phylogenetic tree into segments #################
   times <- node.depth.edgelength(tree) # date of each node
   
@@ -21,7 +21,9 @@ chop_tree <- function(tree) {
   #parent_time[1] <- -99 # placeholder for ancestral state
   
   for (i in 2:length(parent_time)) {
-    parent_time[i] <- (node.depth.edgelength(tree)[node_seq[i]] - node.depth.edgelength(tree)[parent[i]]) / max(node.depth.edgelength(tree))
+    if (scale == T) parent_time[i] <- (node.depth.edgelength(tree)[node_seq[i]] - node.depth.edgelength(tree)[parent[i]]) / max(node.depth.edgelength(tree))
+    
+    if (scale == F)parent_time[i] <- (node.depth.edgelength(tree)[node_seq[i]] - node.depth.edgelength(tree)[parent[i]])
   }
   
   N_seg <- length(node_seq) # total num segments in the tree

### 2022-06-23 chop_tree.R
add stan program for fitting model

@@ -18,7 +18,7 @@ chop_tree <- function(tree, scale=T) {
   
   # Parent time indicates amount of time since the parent node, scaled by the total depth of the tree
   parent_time <- rep(NA, length(node_seq))
-  #parent_time[1] <- -99 # placeholder for ancestral state
+  parent_time[1] <- -99 # placeholder for ancestral state
   
   for (i in 2:length(parent_time)) {
     if (scale == T) parent_time[i] <- (node.depth.edgelength(tree)[node_seq[i]] - node.depth.edgelength(tree)[parent[i]]) / max(node.depth.edgelength(tree))
@@ -28,12 +28,18 @@ chop_tree <- function(tree, scale=T) {
   
   N_seg <- length(node_seq) # total num segments in the tree
   
+  # Indicate whether a node in the seq is a tip
+  N_tips <- length(tree$tip.label)
+  tip <- ifelse(node_seq > N_tips, 0, 1)
+  
   return(
     list(
+      N_tips = N_tips,
       N_seg = N_seg,
       node_seq = node_seq,
       parent = parent,
-      ts = parent_time
+      ts = parent_time,
+      tip = tip
       )
   )
 }

### 2022-06-23 delta_theta.R

@@ -1,0 +1,18 @@
+delta_theta <- function(X, A, b) {
+  
+  J <- ncol(X)
+  med_X <- apply(X, 2, median)
+  mad_X <- apply(X, 2, mad)
+  
+  theta_med <- rep(NA, J)
+  theta_mad <- rep(NA, J)
+  diff_theta <- rep(NA, J)
+  
+  for (j in 1:J) {
+    theta_med[j] = -( sum(med_X[-j]*A[j,-j]) + b[j])/A[j,j]
+    theta_mad[j] = -( sum((med_X[-j] + mad_X[-j])*A[j,-j]) + b[j])/A[j,j]
+    diff_theta[j] = (theta_mad[j] - theta_med[j]) / mad_X[j]
+  }
+  
+  return(diff_theta)
+}

### 2022-06-23 SBC.R
simulation based calibration script

@@ -1,0 +1,71 @@
+library(cmdstanr)
+library(phytools)
+library(phangorn)
+library(posterior)
+source("chop_tree.R")
+
+## Simulate a random phylogenetic tree (pure birth)
+N_tips <- 25
+
+# Vectors to hold rank statistics for A matrix
+A_1_2_rank <- c()
+A_2_1_rank <- c()
+
+# Compile stan models
+sim_mod <- cmdstan_model(stan_file="DPM_sim.stan")
+fit_mod <- cmdstan_model(stan_file="DPM_fit.stan")
+###################################################
+#### Loop over simulations ########################
+N_sims <- 250
+
+for ( s in 1:N_sims ) {
+  
+  tree <- pbtree(n = N_tips)
+  
+  tree_data <- chop_tree(tree)
+  
+  #  Specify number of traits
+  J <- 2
+  
+  tree_data$J <- 2
+  tree_data$resp_type <- c(1,1)
+
+# Draw samples from generative model
+sim <- sim_mod$sample(
+  data=tree_data,
+  iter_warmup = 0,
+  iter_sampling = 1,
+  fixed_param = T,
+  chains = 1
+)
+
+post_sim <- posterior::as_draws_rvars(sim$draws())
+
+A_sim <- draws_of(post_sim$A)[1,,]
+
+# Add simulated outcome data
+tree_data$X <- draws_of(post_sim$X)[1,,]
+
+n_samps <- 250
+n_chains <- 8
+
+fit <- fit_mod$sample(
+  data=tree_data,
+  iter_warmup = n_chains,
+  iter_sampling = 250,
+  init = 0,
+  chains = n_chains,
+  parallel_chains = 8,
+  adapt_delta = 0.9
+)
+
+post_fit <- posterior::as_draws_rvars(fit$draws())
+A_fit <- draws_of(post_fit$A)
+
+A_1_2_rank[s] <- rank( c(A_sim[1,2], A_fit[,1,2]) )[1]
+A_2_1_rank[s] <- rank( c(A_sim[2,1], A_fit[,2,1]) )[1]
+
+} # end loop over simulations
+
+# Save results
+write.csv(data.frame(A_1_2_rank, A_2_1_rank), file="SBC_ranks.csv")

### 2022-06-23 SBC_rstan.R
rstan version of SBC

@@ -1,0 +1,57 @@
+library(rethinking)
+library(phytools)
+library(phangorn)
+source("chop_tree.R")
+
+## Simulate a random phylogenetic tree (pure birth)
+N_tips <- 25
+
+# Vectors to hold rank statistics for A matrix
+A_1_2_rank <- c()
+A_2_1_rank <- c()
+
+# Compile stan models
+sim_mod <- stan_model(file="DPM_sim.stan")
+fit_mod <- stan_model(file="DPM_fit.stan")
+###################################################
+#### Loop over simulations ########################
+N_sims <- 250
+
+for ( s in 1:N_sims ) {
+  
+  tree <- pbtree(n = N_tips)
+  
+  tree_data <- chop_tree(tree)
+  
+  #  Specify number of traits
+  J <- 2
+  
+  tree_data$J <- 2
+  tree_data$resp_type <- c(1,1)
+
+# Draw samples from generative model
+sim <- sampling(sim_mod, data=tree_data, iter=1, chains=1, algorithm="Fixed_param")
+
+post_sim <- extract.samples(sim)
+
+A_sim <- post_sim$A[1,,]
+
+# Add simulated outcome data
+tree_data$X <- post_sim$X[1,,]
+
+n_warmup <- 100
+n_samps <- 250
+n_chains <- 8
+
+fit <- sampling(fit_mod, data=tree_data, warmup=n_warmup, iter=n_samps, chains=8, cores=8, control=list(adapt_delta=0.9), init="0")
+
+post_fit <- extract.samples(fit)
+A_fit <- post_fit$A
+
+A_1_2_rank[s] <- rank( c(A_sim[1,2], A_fit[,1,2]) )[1]
+A_2_1_rank[s] <- rank( c(A_sim[2,1], A_fit[,2,1]) )[1]
+
+} # end loop over simulations
+
+# Save results
+write.csv(data.frame(A_1_2_rank, A_2_1_rank), file="SBC_ranks.csv")

### 2022-06-24 SBC_rstan.R
increase number of SBC sims

@@ -15,7 +15,7 @@ sim_mod <- stan_model(file="DPM_sim.stan")
 fit_mod <- stan_model(file="DPM_fit.stan")
 ###################################################
 #### Loop over simulations ########################
-N_sims <- 250
+N_sims <- 1000
 
 for ( s in 1:N_sims ) {
   
@@ -29,29 +29,33 @@ for ( s in 1:N_sims ) {
   tree_data$J <- 2
   tree_data$resp_type <- c(1,1)
 
-# Draw samples from generative model
-sim <- sampling(sim_mod, data=tree_data, iter=1, chains=1, algorithm="Fixed_param")
-
-post_sim <- extract.samples(sim)
-
-A_sim <- post_sim$A[1,,]
-
-# Add simulated outcome data
-tree_data$X <- post_sim$X[1,,]
-
-n_warmup <- 100
-n_samps <- 250
-n_chains <- 8
-
-fit <- sampling(fit_mod, data=tree_data, warmup=n_warmup, iter=n_samps, chains=8, cores=8, control=list(adapt_delta=0.9), init="0")
-
-post_fit <- extract.samples(fit)
-A_fit <- post_fit$A
-
-A_1_2_rank[s] <- rank( c(A_sim[1,2], A_fit[,1,2]) )[1]
-A_2_1_rank[s] <- rank( c(A_sim[2,1], A_fit[,2,1]) )[1]
+  # Draw samples from generative model
+  sim <- sampling(sim_mod, data=tree_data, iter=1, chains=1, algorithm="Fixed_param")
+  
+  post_sim <- extract.samples(sim)
+  
+  A_sim <- post_sim$A[1,,]
+  
+  # Add simulated outcome data
+  tree_data$X <- post_sim$X[1,,]
+  
+  n_warmup <- 100
+  n_samps <- 350
+  n_chains <- 8
+  
+  fit <- sampling(fit_mod, data=tree_data, warmup=n_warmup, iter=n_samps, chains=8, cores=8, control=list(adapt_delta=0.9), init=0.1)
+  
+  post_fit <- extract.samples(fit)
+  A_fit <- post_fit$A
+  
+  A_1_2_rank[s] <- rank( c(A_sim[1,2], A_fit[,1,2]) )[1]
+  A_2_1_rank[s] <- rank( c(A_sim[2,1], A_fit[,2,1]) )[1]
 
 } # end loop over simulations
 
 # Save results
[…]

### 2022-06-27 SBC_rstan.R
simulate in generated quantities block SBC

@@ -58,4 +58,7 @@ write.csv(data.frame(A_1_2_rank, A_2_1_rank), file="SBC_ranks.csv")
 
 #ranks <- read.csv("SBC_ranks.csv")
 #off_diag_ranks <- cbind(ranks$A_1_2_rank, ranks$A_2_1_rank)
-#hist(off_diag_ranks/1200, breaks=100, main="", xlab="Rank (prop max)", col="skyblue", border=NA)
+#hist(off_diag_ranks/2000, breaks=100, main="", xlab="Rank (prop max)", col="skyblue", border=NA)
+
+#hist(off_diag_ranks[,1]/2001, breaks=30, main="", xlab="Rank (prop max)", col="skyblue", border=NA)
+

### 2022-06-27 manuscript.qmd
a little intro

@@
 ---
-title: "Causal inference for trait co-evolution using dynamic phylogenetic models"
+title: "Causal inference and co-evolutionary contingencies using dynamic phylogenetic models"
 format: html
 editor: visual
+bibliography: references.bib
 ---
@@
 
-Co-evolution can be identified statistically as non-zero between-trait phylogenetic correlations----or any other measure of mutual information over evolutionary time.
-Trait co-evolution is of interest in ecology and evolution when phylogenetic correlations are thought to reflect directed, causal relationships between traits, or when correlated traits share common causes.
+Phylogenetic comparative methods (PCMs) are commonly used to study the co-evolution of traits, spanning topics such as anatomy and physiology [@garland2005phylogenetic; @oconnor2022immunity; @thayer2018impact], life history and behavior [@maclean2012does; @clayton1994comparative; @salguero2016fast; @bielby2007fast; @cornwallis2017cooperation], and cultural evolution [@mace2005phylogenetic; @watts2015broad].
+"Co-evolution" refers to repeatable patterns of trait co-variation over evolutionary time, across species or across populations.
+In practice, co-evolution can be investigated using a diverse family of statistical techniques, depending on the research question and type of data available (Garamszegi, 2014a; Nunn, 2011).
 
-## Running Code
+Trait co-evolution simply describes *how* traits change over time.
+This has value for understanding natural history and can stimulate new research questions.
+However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention, i.e., causality.
+Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations.
 
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+can be identified statistically as non-zero between-trait phylogenetic correlations----or any other measure of mutual information over evolutionary time.
 
-```{r}
-1 + 1
-```
+THREE SCENARIOS TO SIMULATE:
 
-You can add options to executable code like this
+\(1\) Confounding by third variable X_3 -\> X_2, X_3 -\> X_1
 
-```{r}
-#| echo: false
-2 * 2
-```
+\(2\) Correlated drift (CoV\[X_1,X_2\]) != 0)
 
-The `echo: false` option disables the printing of code (only output is displayed).
+\(3\) Including correlated drift (shrinking cross-effects)

### 2022-06-27 manuscript.qmd
Outline popular methods

@@
 Trait co-evolution simply describes *how* traits change over time.
-This has value for understanding natural history and can stimulate new research questions.
+Such descriptions have value for understanding natural history, ancestral state reconstruction (using the present to retrodict the past), and can stimulate new research questions.
 However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention, i.e., causality.
 Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations.
+We briefly review some popular classes of PCM and their limitations with respect to causal inference.
+
+## "Static" PCMs
+
+Independent contrasts, PGLS, PGLMMs, PPA
+
+## "Dynamic" PCMs
+
+Pagel's discrete, Hidden Markov Models, Butler & King OU
 

### 2022-06-28 manuscript.qmd
intro static models

@@
 
+Phylogenetic generalized least squares is a variant of linear regression that accounts for the covariance between species or populations due to shared evolutionary history (likewise for its predecessor, independent contrasts).
+The most commonly assumed model of evolution is Brownian motion, where trait X follows a random walk proportional to time ($t$), such that, for any $t$:
+
+$$ E[X] = 0 $$
+$$ \textrm{Var}[X] = \sigma^2 t $$
+Therefore, the variance for a given species or population $j$ is proportional to total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
+For example, imagine a phylogeny with only $J$ = 3 tip species.
+The last common ancestor of all species is dated to T = 10,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago.
+The implied phylogenetic variance-covariance matrix ($V$) under a Brownian motion model is thus:
+
+$$ V = \sigma^2 \begin{bmatrix}
+T, & 0, & 0 \\
+0, & T, & 0.5T \\
+0, & 0.5T & T
+\end{bmatrix} $$
+
+Thus, the total variance for species 3 is the sum of its evolutionary history, $\sigma^2$[t(1,2,3) - t(2,3)] + $\sigma^2$[t(2,3) - t(3)] and the covariance between species 2 and 3 is equal to the sum of their *shared* evolutionary history $\sigma^2[t(2,3) - t(3)]$, which is always some fraction of the species total evolutionary time $T_j$.
+So long as the tree is ultrametric (all tips equidistant from the root), $T_j$ = $T ~\forall ~j \in J$.
+It is easier to interpret $\sigma$ when we first scale all times by the total tree length $T$.
+We will denote this scaled variance-covariance matrix as $V_s$
+
+$$ V_s = \frac{V}{T} = \sigma^2 \begin{bmatrix}
+1, & 0, & 0 \\
+0, & 1, & 0.5 \\
+0, & 0.5 & 1
+\end{bmatrix} $$
+
+While Brownian motion is the simplest case, in general $V$ is constructed using a matrix
+
+All that is needed is a matrix of phylogenetic distances $D$ (i.e., time since species/populations diverged) and an assumed model of trait evolution (e.g., Brownian motion, Ornstein-Uhlenbeck, Early Burst).
+
+$$
+ Y \sim \mathcal{N}(\boldsymbol{X}\beta, \boldsymbol{\Sigma}) \\
+ \boldsymbol{\Sigma} = \sigma^2(1-D) \\
+ V = \begin{bmatrix}  \end{bmatrix}
+$$
+
+Where $\boldsymbol{X}$ is a design matrix of predictors and $\beta$ is a vector of regression coefficients.
+The strength of the covariance between related species is controlled by the scalar $\lambda \in [0,1]$.
+We scale the distance matrix $D$ by the total length of phylogeny so that taking the complement 1 - $D$ gives us the phylogenetic covariance matrix expected under a Brownian motion model.
+
+and an assumed model of trait evolution (e.g., Brownian motion, Ornstein-Uhlenbeck, Early-Burst)
+
+A sample For example,
+
+Phylogenetic generalized multilevel models (PGLMMs) extend this analytic strategy, accommodating more complex, multivariate model designs [@ives2010phylogenetic; @ives2011generalized].
+While the terminology is not always consistent in the literature, note that "phylogenetic correlation" refers to a .
+A residual correlation refers to the remaining trait covariance *after* accounting for similarity due to shared evolutionary history.
+The residual correlations derived from PCMs are often interpreted as evidence for convergent evolution, adjusted for statistical non-independence and confounding due to phylogeny.
+
 Independent contrasts, PGLS, PGLMMs, PPA

### 2022-06-29 manuscript.qmd
more on models of evolution and PGLS

--- 2022-06-27 manuscript.qmd
@@
 
-Phylogenetic comparative methods (PCMs) are commonly used to study the co-evolution of traits, spanning topics such as anatomy and physiology [@garland2005phylogenetic; @oconnor2022immunity; @thayer2018impact], life history and behavior [@maclean2012does; @clayton1994comparative; @salguero2016fast; @bielby2007fast; @cornwallis2017cooperation], and cultural evolution [@mace2005phylogenetic; @watts2015broad].
+Phylogenetic comparative methods (PCMs) are commonly used to study the co-evolution of organismal traits, spanning topics such as anatomy and physiology [@garland2005phylogenetic; @oconnor2022immunity; @thayer2018impact; @navalon2019evolutionary; @dunn2015evolutionary], life history and behavior [@maclean2012does; @clayton1994comparative; @salguero2016fast; @bielby2007fast; @cornwallis2017cooperation; @wilkinson2019recurrent], and cultural evolution [@mace2005phylogenetic; @watts2015broad; @navarrete2016coevolution].
 "Co-evolution" refers to repeatable patterns of trait co-variation over evolutionary time, across species or across populations.
@@
 
+### Phylogenetic Generalized Least Squares (PGLS)
+
+Phylogenetic generalized least squares (PGLS) is a variant of linear regression that accounts for the covariance between species or populations due to shared evolutionary history (likewise for its predecessor, independent contrasts) [@grafen1989phylogenetic; @symonds2014primer; @blomberg2012independent].
+PGLS is defined as:
+
+$$Y \sim \mathcal{N}(\boldsymbol{X}\beta, \boldsymbol{\Sigma})$$
+$$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$
+
+Where $\boldsymbol{Y}$ is the outcome, $\boldsymbol{X}$ is a design matrix of predictors, $\beta$ is a vector of regression coefficients, and $\sigma^2$ is the variance.
+The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix $\boldsymbol{V}$.
+To construct $\boldsymbol{V}$, we need (1) a matrix of patristic distances between species (i.e., time since divergence from last common ancestor) and (2) an assumed model of trait evolution.
+
+The most commonly assumed model is Brownian motion, where trait X follows a random walk proportional to time ($t$), such that, for any $t$:
+
+$$ E[X] = 0 $$
+$$ \textrm{Var}[X] = \sigma^2 t $$
+Therefore, the variance for a given species or population $j$ is proportional to total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
+For example, imagine a phylogeny with only $J$ = 3 tip species.
+The last common ancestor of all species is dated to T = 10,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago.
+The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:
+
+$$ \boldsymbol{V} = \sigma^2 \begin{bmatrix}
+T, & 0, & 0 \\
+0, & T, & 0.5T \\
+0, & 0.5T, & T
+\end{bmatrix} $$
+
+Thus, the total variance for species 3 is the sum of its evolutionary history, $\sigma^2[t(2,3)] + \sigma^2[t(3) - t(2,3)]$ and the covariance between species 2 and 3 is equal to the sum of their *shared* evolutionary history $\sigma^2[t(2,3)]$ (the tree time for their most recent common ancestor), which is always some fraction of the species total evolutionary time $T_j$.
+If the tree is ultrametric (all tips equidistant from the root), $T_j$ = $T ~\forall ~j \in J$.
+
+$\sigma$ is easier to interpret when we first scale all times by the total tree length $T$. We will denote such a scaled variance-covariance matrix as $\boldsymbol{V}_s$:
+
+$$ \boldsymbol{V}_s = \frac{\boldsymbol{V}}{T} = \sigma^2 \begin{bmatrix}
+1, & 0, & 0 \\
+0, & 1, & 0.5 \\
+0, & 0.5, & 1
+\end{bmatrix} $$
+
+Empirically, the actual degree of covariance between related species or populations may be less than expected under a pure Brownian motion model.
+For example, when evolution is rapid (or more mundanely, when measurement error on trait data is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic].
+PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$.
+For example:
+
+$$ \boldsymbol{V_s} = \sigma^2 \begin{bmatrix}
+1, & 0, & 0 \\
+0, & 1, & \lambda0.5 \\
+0, & \lambda0.5, & 1
+\end{bmatrix} $$
+
+This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "squishing" the past co-evolutionary branches.
+$\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history.
+At the extreme of $\lambda = 0$, $\boldsymbol{V}_s$ becomes an identity matrix and PGLS reduces to standard least squares regression.
+
+Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree.
+Other popular choices include Ornstein-Uhlenbeck, where change due is due to both random drift and "stabilizing selection" towards some optimal trait value, and Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time.
+In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.
+
+PGLMMs $\lambda$ is equivalent to R^2.
+
+### Phylogenetinc Generalized Multilevel Models
+
+Phylogenetic generalized multilevel models (PGLMMs) extend this analytic strategy, accommodating more complex, multivariate model designs [@ives2010phylogenetic; @ives2011generalized].
+While the terminology is not always consistent in the literature, note that "phylogenetic correlation" refers to a .
+A residual correlation refers to the remaining trait covariance *after* accounting for similarity due to shared evolutionary history.
+The residual correlations derived from PCMs are often interpreted as evidence for convergent evolution, adjusted for statistical non-independence and confounding due to phylogeny.
+
 Independent contrasts, PGLS, PGLMMs, PPA
+
+Discrete data can be modeled in the GLMMM framework if we conceptualize discrete phenotypes (binary, ordinal, categorical) as manifestations of a latent continuous trait, as in the quantitative genetic threshold model [@wright1934; @felsenstein2005using].
 

### 2022-06-29 manuscript.qmd
onto PGLMMs

@@
 "Co-evolution" refers to repeatable patterns of trait co-variation over evolutionary time, across species or across populations.
-In practice, co-evolution can be investigated using a diverse family of statistical techniques, depending on the research question and type of data available (Garamszegi, 2014a; Nunn, 2011).
+In practice, co-evolution can be investigated using a diverse family of statistical techniques, depending on the research question and type of data available [@harvey1991comparative; @garamszegi2014modern; @nunn2011comparative].
 
@@
 ## "Static" PCMs
+
+Static PCMs quantify co-evolution between traits using the evolutionary regression or correlation coefficient [@pagel1993seeking].
+The evolutionary regression coefficient is an undirected association between traits over evolutionary time is $b = \frac{\sigma(X,Y)}{\sigma^2(X)}$, and the evolutionary correlation coefficient is the standardized coefficient $\rho = \frac{\sigma(X,Y)}{\sigma(X)\sigma(Y)}$ which can be interpreted as "a one standard deviation increase in trait $X$ is associated with a $\rho$ standard deviation change in trait $Y$."
+
+PHYLOGENETIC POSITION CAN BE THOUGHT OF AS A CONTINOUS CATEGORICAL RANDOM VARIABLE
 
@@
 Where $\boldsymbol{Y}$ is the outcome, $\boldsymbol{X}$ is a design matrix of predictors, $\beta$ is a vector of regression coefficients, and $\sigma^2$ is the variance.
+$\mathcal{N}$ denotes a normal or Gaussian probability density function.
 The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix $\boldsymbol{V}$.
@@
 
-The most commonly assumed model is Brownian motion, where trait X follows a random walk proportional to time ($t$), such that, for any $t$:
+The most commonly assumed model is Brownian motion (also called the Wiener process and thus denoted $W$), where $W$ follows a random walk proportional to time ($T$), such that, for any time $t$:
 
-$$ E[X] = 0 $$
-$$ \textrm{Var}[X] = \sigma^2 t $$
+$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$
 Therefore, the variance for a given species or population $j$ is proportional to total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
@@
 For example, when evolution is rapid (or more mundanely, when measurement error on trait data is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic].
-PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$.
+PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$ [@pagel1999inferring].
 For example:
@@
 
-This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "squishing" the past co-evolutionary branches.
+This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches.
 $\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history.
@@
 
-PGLMMs $\lambda$ is equivalent to R^2.
+To recap, $\beta$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species.
+Although PGLS is widely used in comparative biology, it has key limitations such as: assumption of Gaussian errors, prone to overfitting due to lack of parameter regularization, and cannot accommodate many common data structures including repeated measures of species trait values, missing data, and measurement error.
+In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.
 
-### Phylogenetinc Generalized Multilevel Models
+### Phylogenetic Generalized Multilevel Models
 
 Phylogenetic generalized multilevel models (PGLMMs) extend this analytic strategy, accommodating more complex, multivariate model designs [@ives2010phylogenetic; @ives2011generalized].
+
 While the terminology is not always consistent in the literature, note that "phylogenetic correlation" refers to a .

### 2022-06-30 manuscript.qmd
more on causal inference

@@
 
-PHYLOGENETIC POSITION CAN BE THOUGHT OF AS A CONTINOUS CATEGORICAL RANDOM VARIABLE
+#### When can be interpret the evolutionary regression as a causal effect?
+
+```{r, message=FALSE, warning=FALSE}
+require(rethinking)
+require(dagitty)
+
+phy_no_confound <- dagitty( "dag {
+                            PHY -> Y
+                            X -> Y
+                            }")
+
+phy_confound <- dagitty( "dag {
+                            PHY -> Y
+                            PHY -> X
+                            X -> Y
+                            }")
+
+par(mfrow=c(1,2))
+drawdag(phy_no_confound)
+drawdag(phy_confound)
+
+```
+To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means.
+Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$.
+The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
+If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
+Note here that $X$ refers to a random variable while $x$ refers to some unit value of that variable.
 
@@
 Phylogenetic generalized least squares (PGLS) is a variant of linear regression that accounts for the covariance between species or populations due to shared evolutionary history (likewise for its predecessor, independent contrasts) [@grafen1989phylogenetic; @symonds2014primer; @blomberg2012independent].
-PGLS is defined as:
+Given some outcome $Y$ and a co-evolutionary predictor $X$, an observation $i$ from species or population $j$ is modeled as:
 
-$$Y \sim \mathcal{N}(\boldsymbol{X}\beta, \boldsymbol{\Sigma})$$
+$$y_{ij} = \beta_0 + \beta_xx_{ij} + \epsilon_{ij}$$
+$$ \epsilon \sim  \mathcal{N}(0, \boldsymbol{\Sigma})$$
 $$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$
-
-Where $\boldsymbol{Y}$ is the outcome, $\boldsymbol{X}$ is a design matrix of predictors, $\beta$ is a vector of regression coefficients, and $\sigma^2$ is the variance.
-$\mathcal{N}$ denotes a normal or Gaussian probability density function.
+Where $\mathcal{N}$ denotes a normal or Gaussian probability density function and $\sigma^2$ is the variance of $Y$ conditional on $X$.
+$\beta_x$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species.
 The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix $\boldsymbol{V}$.
@@
 
-This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches.
+This scaling is often referred to a≠s a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches.
 $\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history.
@@
 
-To recap, $\beta$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species.
 Although PGLS is widely used in comparative biology, it has key limitations such as: assumption of Gaussian errors, prone to overfitting due to lack of parameter regularization, and cannot accommodate many common data structures including repeated measures of species trait values, missing data, and measurement error.
@@
 
-Phylogenetic generalized multilevel models (PGLMMs) extend this analytic strategy, accommodating more complex, multivariate model designs [@ives2010phylogenetic; @ives2011generalized].
+Phylogenetic generalized multilevel models (PGLMMs) extend the study of trait coevolution, leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized].
+Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' and 'varying' effects) to total trait variance.
+For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:
+
+$$ y_{i,j} = \beta_0 + \beta_{x} x_{ij} + \nu_{j} + \epsilon_{ij} \\$$
+$$ \nu \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\nu}) \\$$
+$$\epsilon_{ij} \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\epsilon}) \\$$
+$$\boldsymbol{\Sigma}_{\nu} = \sigma^2_{\nu}\boldsymbol{V}_s \\$$
+$$\boldsymbol{\Sigma}_{\epsilon} = \sigma^2_{\epsilon}\boldsymbol{I} \\$$
+
+Where $\boldsymbol{I}$ is an identity matrix and thus the residuals $\epsilon$ are assumed conditionally independent.
+The "phylogenetic signal" is now estimated as $\lambda_{\textrm{PGLMM}} = \frac{\sigma_{\nu}^2}{\sigma_{\nu}^2 + \sigma_{\epsilon}^2}$ (adjusted for $X$) and equivalent to $\lambda_{\textrm{PGLS}}$, but with the greater flexibility afforded by PGLMMs.
+$\lambda_{\textrm{PGLMM}}$ is also sometimes referred to as "phylogenetic heritability" [@hadfield2010general].
+
+In this model, $\beta_{x(\textrm{PGLMM})}$ has the same interpretation as $\beta_{x(\textrm{PGLS})}$.
+If the response was non-Gaussian, $\beta_{x(\textrm{PGLS})}$ has the same interpretation *on the latent scale*.
+Conversion of the phylogenetic regression coefficient $\beta_{x(\textrm{PGLMM})}$ to the observation scale requires inverting the link function, and will be sensitive to other model parameters.
+
+#### Phylogenetic Path Analysis & Causal Inference
 

### 2022-07-01 SBC_rstan.R
minor edits and comments on manuscript

@@ -10,6 +10,7 @@ N_tips <- 25
 A_1_2_rank <- c()
 A_2_1_rank <- c()
 
+
 # Compile stan models
 sim_mod <- stan_model(file="DPM_sim.stan")
 fit_mod <- stan_model(file="DPM_fit.stan")

### 2022-07-01 manuscript.qmd

@@
 Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations.
-We briefly review some popular classes of PCM and their limitations with respect to causal inference.
+We briefly review some popular classes of PCM and their limitations with respect to causal inference, before introducing a new method which overcomes some of these limitations.
 
@@
 
-#### When can be interpret the evolutionary regression as a causal effect?
+#### When can we interpret the evolutionary regression as a causal effect?
 
@@
 ```
+
 To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means.
@@
 If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
-Note here that $X$ refers to a random variable while $x$ refers to some unit value of that variable.
+Note here that $X$ refers to some variable while $x$ refers to some unit value of that variable.
 
@@
 
-$$y_{ij} = \beta_0 + \beta_xx_{ij} + \epsilon_{ij}$$
-$$ \epsilon \sim  \mathcal{N}(0, \boldsymbol{\Sigma})$$
-$$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$
-Where $\mathcal{N}$ denotes a normal or Gaussian probability density function and $\sigma^2$ is the variance of $Y$ conditional on $X$.
+$$y_{ij} = \beta_0 + \beta_xx_{ij} + \epsilon_{ij}$$ $$ \epsilon \sim  \mathcal{N}(0, \boldsymbol{\Sigma})$$ $$\boldsymbol{\Sigma} = \sigma^2\boldsymbol{V} $$ Where $\mathcal{N}$ denotes a normal or Gaussian probability density function and $\sigma^2$ is the variance of $Y$ conditional on $X$.
 $\beta_x$ is an estimator of the evolutionary regression coefficient that takes into account the phylogenetic non-independence between tip species.
-The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix $\boldsymbol{V}$.
+The covariance between related species is proportional to the off-diagonals of the phylogenetic variance-covariance matrix <!--# isn't this also often called A?
+doesn't matter i guess..
+--> $\boldsymbol{V}$.
 To construct $\boldsymbol{V}$, we need (1) a matrix of patristic distances between species (i.e., time since divergence from last common ancestor) and (2) an assumed model of trait evolution.
@@
 
-$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$
-Therefore, the variance for a given species or population $j$ is proportional to total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
+$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$ Therefore, the variance for a given species or population $j$ is proportional to the total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
 For example, imagine a phylogeny with only $J$ = 3 tip species.
-The last common ancestor of all species is dated to T = 10,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago.
+The last common ancestor of all species is dated to T = 10,000 years ago <!--# maybe make it 1 million?
+more realistic for species -->, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago.
 The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:
@@
 
-This scaling is often referred to a≠s a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches.
+This scaling is often referred to as a "branch length transformation," because shrinking the off-diagonals (shared evolutionary history) while holding the diagonals constant (independent evolution) is equivalent to "stretching" the terminal branches leading to the tips and "shrinking" the past co-evolutionary branches.
 $\lambda = 1$ indicates that the phylogenetic strength is as strong as expected under a Brownian motion model, while lower values means that a lower proportion of the trait variance can be attributed to shared evolutionary history.
@@
 Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree.
-Other popular choices include Ornstein-Uhlenbeck, where change due is due to both random drift and "stabilizing selection" towards some optimal trait value, and Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time.
+Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time.
 In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.
 
-Although PGLS is widely used in comparative biology, it has key limitations such as: assumption of Gaussian errors, prone to overfitting due to lack of parameter regularization, and cannot accommodate many common data structures including repeated measures of species trait values, missing data, and measurement error.
+Although PGLS is widely used in comparative biology, it has several key limitations.
+These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data structures such as repeated measures of species trait values, missing data, and measurement error.
 In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.
@@
 
-Phylogenetic generalized multilevel models (PGLMMs) extend the study of trait coevolution, leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized].
-Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' and 'varying' effects) to total trait variance.
+Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->.
+Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance.
 For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:
 
-$$ y_{i,j} = \beta_0 + \beta_{x} x_{ij} + \nu_{j} + \epsilon_{ij} \\$$
-$$ \nu \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\nu}) \\$$
-$$\epsilon_{ij} \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\epsilon}) \\$$
-$$\boldsymbol{\Sigma}_{\nu} = \sigma^2_{\nu}\boldsymbol{V}_s \\$$
-$$\boldsymbol{\Sigma}_{\epsilon} = \sigma^2_{\epsilon}\boldsymbol{I} \\$$
+$$ y_{i,j} = \beta_0 + \beta_{x} x_{ij} + \nu_{j} + \epsilon_{ij} \\$$ $$ \nu \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\nu}) \\$$ $$\epsilon_{ij} \sim \mathcal{N}(0,\boldsymbol{\Sigma}_{\epsilon}) \\$$ $$\boldsymbol{\Sigma}_{\nu} = \sigma^2_{\nu}\boldsymbol{V}_s \\$$ $$\boldsymbol{\Sigma}_{\epsilon} = \sigma^2_{\epsilon}\boldsymbol{I} \\$$
 
@@
 The "phylogenetic signal" is now estimated as $\lambda_{\textrm{PGLMM}} = \frac{\sigma_{\nu}^2}{\sigma_{\nu}^2 + \sigma_{\epsilon}^2}$ (adjusted for $X$) and equivalent to $\lambda_{\textrm{PGLS}}$, but with the greater flexibility afforded by PGLMMs.
-$\lambda_{\textrm{PGLMM}}$ is also sometimes referred to as "phylogenetic heritability" [@hadfield2010general].
+$\lambda_{\textrm{PGLMM}}$ is also sometimes referred to as "phylogenetic heritability" [@hadfield2010general], since it is calculated in the same way as heritability based on pedigrees in quantitative genetics.
 

### 2022-07-07 manuscript.qmd
simulations of phylogenetic confounding

@@
 We briefly review some popular classes of PCM and their limitations with respect to causal inference, before introducing a new method which overcomes some of these limitations.
+
+TWO different reasons to "adjust for phylogeny"
+
+(1) Repeated measures of closely-related species, non-independence (std errors too small)
+
+(2) Clustering in multivariate data can lead to confounding due to sampling variation.
+Type I (between-cluster sampling variation leads to spurious correlation) and Type II error (between cluster sampling variation leads to underestimated correlation).
+
+Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.
 
@@
 The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
-If the probability distribution of $Y$ changes based on the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
-Note here that $X$ refers to some variable while $x$ refers to some unit value of that variable.
+If the probability distribution of $Y$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
+Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable.
 
@@
 
-Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->.
+Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->.
 Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance.
@@
 Conversion of the phylogenetic regression coefficient $\beta_{x(\textrm{PGLMM})}$ to the observation scale requires inverting the link function, and will be sensitive to other model parameters.
+
+#### Multivariate PGLMMs
+
+Residual and phylogenetic correlations can be different!
+Why?
+Plasticity and/or microevolutionary response to X may not be the same as the macroevolutionary process, a violation of *ergodicity*.
+or correlated measurement error.
 

### 2022-07-07 clade_var.R

@@ -1,0 +1,68 @@
+library(wesanderson)
+library(patchwork)
+library(ggplot2)
+
+N <- 400
+N_clades <- 4
+
+clade_cols <- wes_palette("Darjeeling1")
+
+#### First graph = type I error ############
+set.seed(12)
+
+clade <- sample(1:N_clades, size=N, replace=T)
+clade_X <- rnorm(N_clades, 0, 1) * 1
+clade_Y <- rnorm(N_clades, 0, 1) * 1
+  
+X <- rnorm(N, clade_X[clade], 1)
+Y <- rnorm(N, clade_Y[clade], 1)
+
+cor(X,Y)
+
+d <- data.frame(
+  X = scale(X),
+  Y = scale(Y),
+  clade = paste0("Y ~ X | Clade = ", as.character(clade))
+)
+
+type_I_plot <- ggplot(d, aes(x=X, y=Y, color=clade)) +
+  geom_point(alpha=0.6) +
+  geom_smooth(method="lm",aes(fill=clade)) +
+  geom_smooth(method="lm",aes(fill="Y ~ X", color="Y ~ X")) +
+  scale_color_manual(values=c("black", clade_cols )) +
+  scale_fill_manual(values=c("black", clade_cols)) +
+  theme_bw(base_size=12) +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.title = element_blank())
+
+#### Second graph = type II error ############
+set.seed(202)
+#202
+
+clade <- sample(1:N_clades, size=N, replace=T)
+clade_X <- rnorm(N_clades, 0, 1) * 1
+clade_Y <- rnorm(N_clades, 0, 1) * 1
+
+X <- rnorm(N, clade_X[clade], 1)
+Y <- rnorm(N, X*0.5 + clade_Y[clade], 1)
+
+cor(X,Y)
+
+d <- data.frame(
+  X = scale(X),
+  Y = scale(Y),
+  clade = paste0("Y ~ X | Clade = ", as.character(clade))
+)
+
+type_II_plot <- ggplot(d, aes(x=X, y=Y, color=clade)) +
+  geom_point(alpha=0.6) +
+  geom_smooth(method="lm",aes(fill=clade)) +
+  geom_smooth(method="lm",aes(fill="Y ~ X", color="Y ~ X")) +
+  scale_color_manual(values=c("black", clade_cols )) +
+  scale_fill_manual(values=c("black", clade_cols)) +
+  theme_bw(base_size=12) +
[…]

### 2022-07-12 manuscript.qmd
Many notes on causal inference in PCMs

@@
 Such descriptions have value for understanding natural history, ancestral state reconstruction (using the present to retrodict the past), and can stimulate new research questions.
-However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention, i.e., causality.
+However, simply inferring that traits changed together in the past does not tell us *why* they changed, or how the traits would respond to intervention.
+In short, evolutionary correlation does not imply evolutionary causation.
 Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations.
-We briefly review some popular classes of PCM and their limitations with respect to causal inference, before introducing a new method which overcomes some of these limitations.
+While some authors use paradigms such as phylogenetic path analysis to justify their models and make their causal assumptions explicit, many others refrain from stating their assumptions at all, despite drawing implicit causal conclusions.
+Ambiguity about causal inference can lead to misleading claims and contradictory results, undermining the perceived value and reliability of comparative approaches in biology.
 
-TWO different reasons to "adjust for phylogeny"
+A major barrier to incorporating causal inference into study design, analysis, and interpretation is that the causal assumptions baked into popular PCMs are not always apparent to the end users (researchers).
+For example, the most widely used model of trait evolution in PCMs (Brownian motion) assumes a purely neutral model of evolution with a constant rate of drift over the entire phylogenetic tree.
+Another example is that, for virtually all models of the co-evolution of continuous traits, effects can only be interpreted causally if we assume that we have the direction of causality right (X -> Y, not Y -> X), and that there is no reciprocal causation.
 
-(1) Repeated measures of closely-related species, non-independence (std errors too small)
+In this article, we first review some popular classes of PCM and their assumptions and limitations with respect to causal inference.
+We then introduce a new method which overcomes some of these limitations and enables the study of directional trait co-evolution for an arbitrary number of traits that can be used with any type of data (continuous, discrete, ordered).
 
-(2) Clustering in multivariate data can lead to confounding due to sampling variation.
-Type I (between-cluster sampling variation leads to spurious correlation) and Type II error (between cluster sampling variation leads to underestimated correlation).
+## Co-evolution as correlated drift: the evolutionary regression coefficient
+
+The most common way to quantify co-evolution between traits is the evolutionary regression coefficient [@pagel1993seeking].
+The evolutionary regression coefficient is a measure of association between traits over evolutionary time, $b = \frac{\sigma(X,Y)}{\sigma^2(X)}$.
+It can be interpreted as "a one unit increase in trait $X$ is associated with a $b$ unit change in trait $Y$." When can we interpret these statistics as causal effects?
+The short answer is...almost never!
+To explain why, we sketch out assumptions needed to map an evolutionary regression coefficient onto a causal estimand.
+
+First, what do we mean by causal?
+Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $P(Y|do(X = x)) \neq P(Y|do(X = \textrm{not x}))$.
+The do($\dot$) operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
+This is an inclusive definition, where if the probability distribution $P(Y)$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
+Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable.
+In any observational study, we cannot directly access $P(Y|do(X = x))$.
+Instead, we have just have the condition probability of $Y$ given $X$, $P(Y|X)$.
+Pearl and colleagues' "do calculus" is a set of rules that tells us under what conditions $P(Y|do(X)) = P(Y|X)$, or when we can make them equivalent by adjustment for other variables $P(Y|do(X),Z) = P(Y|X,Z)$.
+An *estimand* is a unit specific estimate such as "what would we expect species $i$ trait value $y_i$ to be if $x_i$ took on a given value?".
+Under a particular set of causal and parameteric assumptions, a regression coefficient $b$ can be interpeted as the estimand "what is the difference in the expected (mean) value of $Y$ if $x = 1$ vs $x = 0~$?", or $E[Y|x=1] - E[Y|x=0]$.
+
+An unitiuitive aspect of phylogenetic comparative methods is that cross-sections of species traits are used as an estimate of the *change* in $Y$ in response $X$.
+After all, we have only observed current trait values, not their evolution over time.
+It is unituitive not because it is incorrect in principle, but because it requires the reseracher to make a series of strong assumptions that may be in conflict with the stated research question.
+Following [@pagel1993seeking], we define a causal model of trait co-evolution in which we would be justified in interpreting the evolutionary regression coefficient as a causal effect.
+The model is defined below and depicted graphically in Figure 1.
+
+Given a time series of trait evolution with $T$ time points, $\forall ~t \in T$
+
+$$ x_t = x_{t-1} + \epsilon_{x,t}$$
+$$ y_t = y_{t-1} + b(x_t - x_{t-1}) + \epsilon_{y,t} = y_{t-1} + b\epsilon_{x,t} + \epsilon_{y,t} $$
+$$ \epsilon_{x,t} \sim \mathcal{N}(0,\sigma_x^2) $$
+$$ \epsilon_{y,t} \sim \mathcal{N}(0,\sigma_y^2) $$
+Expressed as discrete time difference equations:
+
+$$ \Delta x_t =  \epsilon_{x,t}$$
+$$ \Delta y_t = b\epsilon_{x,t} + \epsilon_{y,t}$$
+This model assumes that all change in trait $X$ is due to random drift $\epsilon_x$.
+Likewise, change in trait $Y$ is due to independent drift $\epsilon_y$ and response to *changes* in $X$.
+There is no directional change over time, and the evolution of both traits is state-independent (i.e., $Y$ has an invariant response to drift in $X$, regardless of how high or low it is).
+We can relax some assumptions, such as constant $\sigma$ and $\b$ over time, and still get unbiased estimates of $b$ using a cross-sectional regression of species.
+Additionally, if $Y$ has a non-linear response to $\epsilon_x$, then $b$ will still be a valid estimate of the linear component--albeit now possibly misleading if the total response exhibits strong non-linearity.
+But if either trait is under selection (i.e., the edges between $Y_t$ and $Y_{t+1}$ or between $X_t$ and $X_{t+1}$ aren't = 1), then then regression coefficient will no longer be equal to b.
+To get a sense for why, imagine that, in addition to drift, $X$ is under moderate stabilizing selection.
+This means that $X_t$ is no longer just a record of the additive evolutionary history of $\epsilon_x$, and that a causal effect of $X$ on $Y$ is not evident from the trait covariance at the tips of a phylogeny.
+Just because you can't see it anymore, doesn't mean it isn't important.
+
+![Figure 1](figures/MA_DAG.png){width=60%}
+
+We previously stated that a cross section of current species trait values can provide an unbiased estimate of the change in trait $Y$ in response to changes in $X$.
+This fact is well known in the comparative literature [@pagel1993seeking; @rohle2006comment], and consistent with the rules of do-calculus--given the strict assumptions of the model outined before.
+However, if traits were correlated in the past but no longer exhibit any association, then evolutionary history is a confounder.
+For example, in Figure 4B, $U_0$ creates a "backdoor path" that connects all pairs of $X_t$ and $Y_t$.
+
+Phylogentic position is just a type of clustering, and the presence of clusters (such as a clade) in comparative datasets makes analyses more susceptible to Simpson's paradox, a reversal of trend at group (clade) vs individual (species) level.
+Statistical adjustment for phylogeny helps reduce both the type I (false-positive) and type II (false-negative) error rates (plus the Type M and Type S errors).
+It should be noted here that "adjusting for phylogeny" is not a magic-fix for unobserved confounding that is persistent throughout the phylogenetic tree.
+But it can help us adjust for confounding in the past.
+
+HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE.
+
+Additionally, even if there was zero structural phylogenetic confounding, it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$.
+Marginalizing over phylogenetic position can also give us imrpoved estimtes of the average effects when clades are unequally sampled.
+
+This model assumes that trait co-evolution is *ergodic*, which means that the time average of the co-evolutionary process is equal to the ensemble average.
+In other words, a cross-section contains the same information as a time series.
+Remarkably, this means that even a simple species regression (i.e., regression of $Y$ on $X$, with no phylogenetic information considered) provides an unbiased estimate of $b$.
+
+An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works. But if the predictor is under selection, things get messed up.
 
@@
 
-## "Static" PCMs
+OU ADAPATIVE EVOLUTION AREN'T JUST A FIX FOR VIOLATION OF BROWNIAN MOTION--THEY IMPLY AN ENTIRELY DIFFERENT THEORETICAL ESTIMAND!
+Difference in Y given difference in X vs difference in the equilibrium trait value of Y given X.
+In the latter, the mean is not ergodic?
 
-Static PCMs quantify co-evolution between traits using the evolutionary regression or correlation coefficient [@pagel1993seeking].
-The evolutionary regression coefficient is an undirected association between traits over evolutionary time is $b = \frac{\sigma(X,Y)}{\sigma^2(X)}$, and the evolutionary correlation coefficient is the standardized coefficient $\rho = \frac{\sigma(X,Y)}{\sigma(X)\sigma(Y)}$ which can be interpreted as "a one standard deviation increase in trait $X$ is associated with a $\rho$ standard deviation change in trait $Y$."
+STATE DEPENDENT ANALYSES MAKE US MORE SUSCEPTIBLE TO DARWIN'S SCENARIO -- UNREPLICATED BURSTS.
+REMOVAL OF PHYLOGENETIC CONFOUNDING RELIES ON THE COVARIANCE BETWEEN X AND Y DECREASING OVER TIME.
+IF THEY ARE MAINTAINED BY SELECTION WITHIN A CLADE, WE MAY FOOL OURSELVES.
+(!!!).
+SOLUTION = CLADE SPECIFIC RANDOM EFFECTS.
 
-#### When can we interpret the evolutionary regression as a causal effect?
+## Causality on Phylogenetic Trees
 
-```{r, message=FALSE, warning=FALSE}
-require(rethinking)
-require(dagitty)
+TEST YOUR METHOD AGAINST MODEL MISSPECIFICATION--WHAT IF DATA WERE GENERATED UNDER CORRELATED DRIFT?
+AFTER CLARIFYING HOW THESE ARE FUNDAMENTALLY DIFFERENT QUESTIONS/ESTIMANDS, IMPORTANT TO SHOW HOW THEY INFUENCE EACH OTHERS ESTIMATES IF WE IGNORE THEM.
+MY INTUITION = IGNORING SELECTION LEADS TO BIAS, IGNORING DRIFT DOES NOT.
 
-phy_no_confound <- dagitty( "dag {
-                            PHY -> Y
-                            X -> Y
-                            }")
-
-phy_confound <- dagitty( "dag {
-                            PHY -> Y
-                            PHY -> X
-                            X -> Y
-                            }")
-
-par(mfrow=c(1,2))
-drawdag(phy_no_confound)
-drawdag(phy_confound)
-
-```
-
-To understand when the evolutionary regression coefficient can be interpreted as a causal effect, we must first clarify what "causality" means.
-Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $Pr(Y|do(X = x)) \neq Pr(Y|do(X = \textrm{not x}))$.
-The do() operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
-If the probability distribution of $Y$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
-Note here that $X$ refers to a random variable while $x$ refers to some realized unit value of that variable.
+![Figure 1](figures/phylo_DAG.png)
 
@@
 For example, imagine a phylogeny with only $J$ = 3 tip species.
-The last common ancestor of all species is dated to T = 10,000 years ago <!--# maybe make it 1 million?
-more realistic for species -->, and the last common ancestor of 2 and 3 is dated to 0.5T = 5,000 years ago.
+The last common ancestor of all species is dated to T = 1,000,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 500,000 years ago.
 The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:
@@
 
-Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized] <!--# there's also Lynch 1991 METHODS FOR THE ANALYSIS OF COMPARATIVE DATA IN EVOLUTIONARY BIOLOGY, Evolution -->.
+Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized].
 Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance.
@@
 or correlated measurement error.
+
+PHYLOGENETIC CORRELATION DIRECTLY RELATED TO evolutionary regression coefficient $b$.
+
+$$ \rho(X,Y) = \frac{\sigma(X,Y)}{\sigma(x)\sigma(Y)} $$
+$$ b = \rho(X,Y)\frac{\sigma(Y)}{\sigma(X)} $$
 

### 2022-07-12 manuscript.qmd
minor comments and typos

@@
 In short, evolutionary correlation does not imply evolutionary causation.
-Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations.
-While some authors use paradigms such as phylogenetic path analysis to justify their models and make their causal assumptions explicit, many others refrain from stating their assumptions at all, despite drawing implicit causal conclusions.
+Nonetheless, researchers in ecology and evolution are often motivated to study trait co-evolution because they believe these patterns reflect causal relations [CITATION/EXAMPLE HERE?].
+While some authors use paradigms such as phylogenetic path analysis [POINT TO RELEVANT SECTION BELOW?] to justify their models and make their causal assumptions explicit, many others refrain from stating their assumptions at all, despite drawing implicit causal conclusions.
 Ambiguity about causal inference can lead to misleading claims and contradictory results, undermining the perceived value and reliability of comparative approaches in biology.
+[I WOULD MAKE THIS PARAGRAPH SLIGHTLY LESS 'OFFENSIVE', BY 1) STATING THAT THESE ISSUES ARE BY NO MEANS LIMITED TO PCM, AND 2) GIVE SOME EXAMPLES OF PCMS  THAT DO OR DO NOT DRAW CAUSAL CONCLUSIONS.
+OTHERWISE WE RISK UPSETTING READERS WHO EITHER DON'T THINK THIS IS A PROBLEM, OR DON'T LIKE TO BE SINGLED OUT AS PROBLEMATIC - e.g.
+ISLER & VAN SCHAIK EVO ANTHRO IS A GOOD EXAMPLE OF BEING COGNESCENT OF THE DIFFICULTY OF CAUSAL INFERENCE IN A PGLS/IC WORLD]
 
@@
 In this article, we first review some popular classes of PCM and their assumptions and limitations with respect to causal inference.
-We then introduce a new method which overcomes some of these limitations and enables the study of directional trait co-evolution for an arbitrary number of traits that can be used with any type of data (continuous, discrete, ordered).
+These range from standard regression models that lack clear causal interpretations over phylogenetic path analysis, which makes causal assumptions explicit, to dynamic models that detect temporal contingency (but are limited to binary traits).
+We then introduce a new dynamic method which overcomes some of these limitations and enables the study of directional trait co-evolution for an arbitrary number of traits that can be used with any type of data (continuous, discrete, ordered).
 
@@
 Additionally, if $Y$ has a non-linear response to $\epsilon_x$, then $b$ will still be a valid estimate of the linear component--albeit now possibly misleading if the total response exhibits strong non-linearity.
-But if either trait is under selection (i.e., the edges between $Y_t$ and $Y_{t+1}$ or between $X_t$ and $X_{t+1}$ aren't = 1), then then regression coefficient will no longer be equal to b.
+But if either trait is under selection (i.e., the edges between $Y_t$ and $Y_{t+1}$ or between $X_t$ and $X_{t+1}$ aren't = 1), then the regression coefficient will no longer be equal to b.
 To get a sense for why, imagine that, in addition to drift, $X$ is under moderate stabilizing selection.
@@
 
-Phylogentic position is just a type of clustering, and the presence of clusters (such as a clade) in comparative datasets makes analyses more susceptible to Simpson's paradox, a reversal of trend at group (clade) vs individual (species) level.
-Statistical adjustment for phylogeny helps reduce both the type I (false-positive) and type II (false-negative) error rates (plus the Type M and Type S errors).
+Phylogentic position is just a type of clustering, and the presence of clusters (such as a clade) in comparative datasets makes analyses more susceptible to Simpson's paradox, a reversal of a trend at the group level (clade) vs the individual level (species).
+Statistical adjustment for phylogeny helps reduce both the type I (false-positive) and type II (false-negative) error rates (plus the Type M and Type S errors [ADD CITATION FOR WHAT THESE ARE]).
 It should be noted here that "adjusting for phylogeny" is not a magic-fix for unobserved confounding that is persistent throughout the phylogenetic tree.
@@
 Additionally, even if there was zero structural phylogenetic confounding, it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$.
-Marginalizing over phylogenetic position can also give us imrpoved estimtes of the average effects when clades are unequally sampled.
+Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.
 

### 2022-07-13 manuscript.qmd
round off phyo confounding section

@@
 Following Pearl (some citations), we say that $X$ is a cause of $Y$ when $P(Y|do(X = x)) \neq P(Y|do(X = \textrm{not x}))$.
-The do($\dot$) operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
+The do($\cdot$) operator represents an *intervention* in which the value of some variable is fixed (for example, through assignment in a randomized experiment).
 This is an inclusive definition, where if the probability distribution $P(Y)$ is sensitive to the value of $X$ that is assigned, then $X$ has a causal effect on $Y$.
@@
 An *estimand* is a unit specific estimate such as "what would we expect species $i$ trait value $y_i$ to be if $x_i$ took on a given value?".
-Under a particular set of causal and parameteric assumptions, a regression coefficient $b$ can be interpeted as the estimand "what is the difference in the expected (mean) value of $Y$ if $x = 1$ vs $x = 0~$?", or $E[Y|x=1] - E[Y|x=0]$.
+Under a particular set of causal and parameteric assumptions, a regression coefficient $b$ can be interpeted as the estimand "what is the difference in the expected (mean) value of $Y$ if $x = 1$ vs $x = 0~$?", or $\mathbb{E}[Y|x=1] - \mathbb{E}[Y|x=0]$.
 
@@
 We previously stated that a cross section of current species trait values can provide an unbiased estimate of the change in trait $Y$ in response to changes in $X$.
-This fact is well known in the comparative literature [@pagel1993seeking; @rohle2006comment], and consistent with the rules of do-calculus--given the strict assumptions of the model outined before.
-However, if traits were correlated in the past but no longer exhibit any association, then evolutionary history is a confounder.
+This fact is well known in the comparative literature [@pagel1993seeking; @rohle2006comment], and consistent with the rules of do-calculus--given the strict assumptions of the model outlined before.
+However, if traits were correlated in the past but no longer exhibit any association, then evolutionary history acts as a confounder.
 For example, in Figure 4B, $U_0$ creates a "backdoor path" that connects all pairs of $X_t$ and $Y_t$.
+This means that $P(Y|do(X)) \neq P(Y|X)$, unless we can somehow adjust for $U_0$.
+
+The confounder denoted $U$ is unobserved, meaning that we can't statistically adjust for it.
+However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1}$.
+In other words, the present is conditionally independent of $U_0$ (or "d-seperated") given the past.
+But in most comparative studies, we have not observed the past, only the present.
+This is where phylogeny can help us, because more closely related species or populations have more similar confounding factors (e.g., shared genetic architecture, environmental similarity), on average, than distantly related ones.
+Assuming that we have an appropriate model of evolution, then the patristic distance between species (i.e., time since divergence from a common ancestor) allows us to probabalistically reconstruct unobserved past values of $X$ and $Y$, closing the backdoor path $X \leftarrow U \rightarrow Y$ (cite Rethinking 2nd edition).
+It should be noted that such a scenario is sometimes called "phylogenetic confounding," but...[DOESN"T WORK IF U IS STILL PRESENT AT THE TIPS.
+HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP?
+IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME.
+IN THE LATTER IT IS CONSTANT/STABLE].
 
@@
 
-HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE.
-
-Additionally, even if there was zero structural phylogenetic confounding, it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$.
+Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$.
 Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.
 
-This model assumes that trait co-evolution is *ergodic*, which means that the time average of the co-evolutionary process is equal to the ensemble average.
-In other words, a cross-section contains the same information as a time series.
-Remarkably, this means that even a simple species regression (i.e., regression of $Y$ on $X$, with no phylogenetic information considered) provides an unbiased estimate of $b$.
-
-An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works. But if the predictor is under selection, things get messed up.
-
-Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.
-
-OU ADAPATIVE EVOLUTION AREN'T JUST A FIX FOR VIOLATION OF BROWNIAN MOTION--THEY IMPLY AN ENTIRELY DIFFERENT THEORETICAL ESTIMAND!
-Difference in Y given difference in X vs difference in the equilibrium trait value of Y given X.
-In the latter, the mean is not ergodic?
-
-STATE DEPENDENT ANALYSES MAKE US MORE SUSCEPTIBLE TO DARWIN'S SCENARIO -- UNREPLICATED BURSTS.
-REMOVAL OF PHYLOGENETIC CONFOUNDING RELIES ON THE COVARIANCE BETWEEN X AND Y DECREASING OVER TIME.
-IF THEY ARE MAINTAINED BY SELECTION WITHIN A CLADE, WE MAY FOOL OURSELVES.
-(!!!).
-SOLUTION = CLADE SPECIFIC RANDOM EFFECTS.
-
-## Causality on Phylogenetic Trees
-
-TEST YOUR METHOD AGAINST MODEL MISSPECIFICATION--WHAT IF DATA WERE GENERATED UNDER CORRELATED DRIFT?
-AFTER CLARIFYING HOW THESE ARE FUNDAMENTALLY DIFFERENT QUESTIONS/ESTIMANDS, IMPORTANT TO SHOW HOW THEY INFUENCE EACH OTHERS ESTIMATES IF WE IGNORE THEM.
-MY INTUITION = IGNORING SELECTION LEADS TO BIAS, IGNORING DRIFT DOES NOT.
-
-![Figure 1](figures/phylo_DAG.png)
+Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient and introduce some variations on the causal model in Figure 1.
 
@@
 Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time.
+NOTE ON OU COVARIANCE: An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works.
+But if the predictor is under selection, things get messed up.
 In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.
@@
 ### Phylogenetic Generalized Multilevel Models
+
+Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.
 
@@
 
-#### Multivariate PGLMMs
+### Mutli-response PGLMMs: contrasting macro and microevolution
 
@@
 
-#### Phylogenetic Path Analysis & Causal Inference
+### Misc. Notes
+
+OU ADAPATIVE EVOLUTION AREN'T JUST A FIX FOR VIOLATION OF BROWNIAN MOTION--THEY IMPLY AN ENTIRELY DIFFERENT THEORETICAL ESTIMAND!
+Difference in Y given difference in X vs difference in the equilibrium trait value of Y given X.
+In the latter, the mean is not ergodic?
+
+OU BUTLER AND KING ALGO--PIECEWISE CONSTANT SELECTION REGIME.
+I THINK THIS MEANS THAT Y RESPONDS TO WHATEVER TRIAT VALUE X WAS AT THE START OF THE SEGMENT--DOESN'T RESPOND TO CHANGES IN X WITHIN THE SEG.
+
+STATE DEPENDENT ANALYSES MAKE US MORE SUSCEPTIBLE TO DARWIN'S SCENARIO -- UNREPLICATED BURSTS.
+REMOVAL OF PHYLOGENETIC CONFOUNDING RELIES ON THE COVARIANCE BETWEEN X AND Y DECREASING OVER TIME.
+IF THEY ARE MAINTAINED BY SELECTION WITHIN A CLADE, WE MAY FOOL OURSELVES.
+(!!!).
+SOLUTION = CLADE SPECIFIC RANDOM EFFECTS.
+
+TEST YOUR METHOD AGAINST MODEL MISSPECIFICATION--WHAT IF DATA WERE GENERATED UNDER CORRELATED DRIFT?
+AFTER CLARIFYING HOW THESE ARE FUNDAMENTALLY DIFFERENT QUESTIONS/ESTIMANDS, IMPORTANT TO SHOW HOW THEY INFUENCE EACH OTHERS ESTIMATES IF WE IGNORE THEM.
+MY INTUITION = IGNORING SELECTION LEADS TO BIAS, IGNORING DRIFT DOES NOT.
+
+![Figure 1](figures/phylo_DAG.png)
+### Phylogenetic Path Analysis
 
@@
 
-## "Dynamic" PCMs
+## Dynamic state-space models of co-evolution
 

### 2022-07-14 manuscript.qmd
working through limits of PGLMMs vis-a-vis state-dependence

@@
 
-## Introduction
+## 1. Introduction
 
@@
 
-## Co-evolution as correlated drift: the evolutionary regression coefficient
+## 2. Co-evolution as correlated drift: the state-independent evolutionary regression coefficient
 
@@
 The confounder denoted $U$ is unobserved, meaning that we can't statistically adjust for it.
-However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1}$.
-In other words, the present is conditionally independent of $U_0$ (or "d-seperated") given the past.
+However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1} \}$.
+In other words, the present is conditionally independent of $U_0$ ("d-seperated") given the past.
 But in most comparative studies, we have not observed the past, only the present.
@@
 
-Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$.
+Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phylogeny is a "neutral-good control" in that it reduces variation in $Y$, thus improving the precision of our estimates for the causal effect of $X$ [@cinelli2021crash].
 Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.
 
-Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient and introduce some variations on the causal model in Figure 1.
+Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient, and introduce some variations on the causal model in Figure 1.
 
-### Phylogenetic Generalized Least Squares (PGLS)
+### 2.1 Phylogenetic Generalized Least Squares (PGLS)
 
@@
 $$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$ Therefore, the variance for a given species or population $j$ is proportional to the total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history.
+Note that this is a continuous-time analog of the discrete time model presented in Figure 1.
 For example, imagine a phylogeny with only $J$ = 3 tip species.
@@
 Empirically, the actual degree of covariance between related species or populations may be less than expected under a pure Brownian motion model.
-For example, when evolution is rapid (or more mundanely, when measurement error on trait data is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic].
+For example, when evolution is rapid (or more mundanely, when measurement error is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic].
 PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$ [@pagel1999inferring].
@@
 Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree.
-Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time.
-NOTE ON OU COVARIANCE: An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works.
-But if the predictor is under selection, things get messed up.
-In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.
+Given the assumptions laid out in the previous section, the Brownian motion covariance function allows us to appropriately adjust for phylogenetic confounding and non-independence.
+But if traits do not evolve according to a Brownian motion process, we need a different way to construct $V$.
+In general, $\boldsymbol{V}$ is constructed by inputting a matrix of patristic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.
+For example, Early Burst models of adaptive radiation allow the rate of change ($\sigma$) to decrease over time.
+Another popular choice is the Ornstein-Uhlenbeck (OU) covariance function, which follow from a model of trait evolution where change is due to both stochastic "drift" and deterministic "selection." As a result, the relationship between patristic distance and phylogenetic covariance becomes non-linear.
+Note however that use of an OU covariance function in the PGLS framework still assumes zero selection on the predictors.
 
 Although PGLS is widely used in comparative biology, it has several key limitations.
-These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data structures such as repeated measures of species trait values, missing data, and measurement error.
-In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.
+These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data features such as repeated measures of species trait values, missing data, and measurement error.
+In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations that distinguishes between macro and micro evolutionary correlations.
 
-### Phylogenetic Generalized Multilevel Models
+### 2.2 Phylogenetic Generalized Multilevel Models (PGLMMs)
 
-Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.
-
-Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized].
+Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized].
 Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance.
@@
 Conversion of the phylogenetic regression coefficient $\beta_{x(\textrm{PGLMM})}$ to the observation scale requires inverting the link function, and will be sensitive to other model parameters.
+Modelling discrete data in the PGLMMM framework can be biologically motivated if we conceptualize discrete phenotypes (binary, ordinal, categorical) as manifestations of a latent continuous trait, as in the quantitative genetic threshold model [@wright1934analysis; @felsenstein2005using].
 
-### Mutli-response PGLMMs: contrasting macro and microevolution
+In addition to accommodating non-Gaussian traits, PGLMMs can readily account for measurement error, within-species trait variation, phylogenetic uncertainty, and missing data [@harmon2005effect; @garamszegi2014uncertainties; @jaeggi2020variable].
+Under the PGLMM umbrella, authors have also extended evolutionary models of trait change to include variable rates of drift and directional change across the phylogenetic tree [@revell2021variable; @pagel2022general].
+
+### 2.3 Mutlivariate PGLMMs: partitioning macro and microevolution
 
@@
 
+While the terminology is not always consistent in the literature, note that "phylogenetic correlation" refers to a .
+A residual correlation refers to the remaining trait covariance *after* accounting for similarity due to shared evolutionary history.
+The residual correlations derived from PCMs are often interpreted as evidence for convergent evolution, adjusted for statistical non-independence and confounding due to phylogeny.
+
 PHYLOGENETIC CORRELATION DIRECTLY RELATED TO evolutionary regression coefficient $b$.
@@
 $$ b = \rho(X,Y)\frac{\sigma(Y)}{\sigma(X)} $$
+The variance partioning approach can be useful, but to study adaptive co-evolution we will need to move beyond the GLMM world, utilizing state-space models.
 
@@
 ![Figure 1](figures/phylo_DAG.png)
-### Phylogenetic Path Analysis
+### 2.4 Phylogenetic Path Analysis
 
-While the terminology is not always consistent in the literature, note that "phylogenetic correlation" refers to a .
-A residual correlation refers to the remaining trait covariance *after* accounting for similarity due to shared evolutionary history.
-The residual correlations derived from PCMs are often interpreted as evidence for convergent evolution, adjusted for statistical non-independence and confounding due to phylogeny.
+## 3. Dynamic state-space models of co-evolution
 
-Independent contrasts, PGLS, PGLMMs, PPA
+### 3.1 Ornstein-Uhlenbeck and co
 
-Discrete data can be modeled in the GLMMM framework if we conceptualize discrete phenotypes (binary, ordinal, categorical) as manifestations of a latent continuous trait, as in the quantitative genetic threshold model [@wright1934; @felsenstein2005using].
+### 3.2 Discrete, HMM
 
-## Dynamic state-space models of co-evolution
+Pagel's classic limited to two binary traits, diveristree and HMM add flexibility to more than two states and variable transition rates throughout the tree, but remain limited in the discrete setting.
+Given infinite data, discrete transition matrix is good, but in finite samples it is problematic....
+Unfortunately though this method is fundamentally limited to discrete traits, sometimes forcing researchers to dichotomize continuous variables (e.g.
+environmental harshness [@cornwallis2017cooperation]), which can lead to loss of power and inconsistent results [@dawson2012dichotomizing; @royston2006dichotomizing].
+Because of this limitation to binary traits this method has not been applied to topics typically involving continuous traits (e.g.
+morphology, life history, brain size, etc.).
 
-Pagel's discrete, Hidden Markov Models, Butler & King OU
-
-can be identified statistically as non-zero between-trait phylogenetic correlations----or any other measure of mutual information over evolutionary time.
+### 3.3 GDPM: how our method combines features of OU and Discrete, and PGLMMs
 
@@
 \(3\) Including correlated drift (shrinking cross-effects)
+
+IDEA: In simulation study, compare statistical power (PP > 0.9, or whatever) vs phylogenetic signal.
+
+## 4. Case study: Primate life history and niche complexity

### 2022-07-15 manuscript.qmd
Update manuscript.qmd

@@
 
-### 3.3 GDPM: how our method combines features of OU and Discrete, and PGLMMs
+## 4. GDPM: A new method that combines features of OU and Discrete, and PGLMMs
 
@@
 
-## 4. Case study: Primate life history and niche complexity
+## 5. Case study: Primate brain size, life history and socio-ecological niche
+
+### 5.1. Background and anaylsis approach
+To illustrate the potential for novel contributions of our method with an empirical example we focus on the co-evolution of brain size, life history, and socio-ecological factors in primates.
+There is a rich literature trying to explain variation in brain size in primates (and other taxa), motivated in part by the fact that we humans have exceptionally large brains for our body size [@smaers2021evolution; @miller2019quantitative; @dunbar2017there; @powell2017re; @isler2014humans].
+Most studies have related variation in brain size to potential predictors using PGLS (or independent contrasts, see 2.1.), with some more recent ones using phylogenetic path analysis [@navarrete2016coevolution] (see 2.4.).
+None, to our knowledge, have used dynamic causal inference though as Pagel's discrete method (see 3.2.) does not lend itself to continuous variables like brain size.
+The literature recognizes the difficulty of causal inference from these largely correlational analyses, stating for example that larger brains could just as well be a cause or a consequence of improved diet [@isler2014humans].
+Furthermore, many predictors of brain size are highly collinear and missing data are common, resulting in different inferences depending on the specifics of the sample [@powell2017re; @wartel2019whatever].
+There is therefore an urgent need for methods that can disentangle leaders from followers and detect reciprocal causation in the evolution of brain size, while flexibly accounting for collinearity, missing data and other issues like measurement error [@sandel2016assessing; @garamszegi2014uncertainties] or clade-specific allometry [@smaers2021evolution].
+
+We used a dataset of x species from [@decasien2017primate], which used PGLS models predicting brain size as a function of body size, diet (degree of frugivory) and sociality (group size).
+We augmented this with life-history variables (female age at maturity, longevity) from [@herculano2019longevity] to capture the cost of brain expansion in terms of delayed reproduction [@isler2009expensive], and downloaded a consensus phylogeny from 10ktrees [@arnold201010ktrees].
+Rather than treating all these variables as independent causes of brain size, which is the implicit assumption of a standard multivariate PGLS, we inferred a plausible causal structure using exploratory graph analysis (EGA) [@martin2019ega+], which detects latent causes; a more comprehensive approach would compare this to causal structures derived from the literature and appropriate benchmarks [@martin2019ega+; @ringen2021novel].
+Instead of modeling absolute brain size and including body size as a covariate, we directly modeled the allometric slope of the brain-body relationship.
+We also accounted for clade-specific allometry [@smaers2021evolution] using species-level intercepts and slopes and imputed missing data rather than dropping incomplete cases.
+For more detail see the accompying R code.
+
+### 5.2. Results and interpretation
+The EGA inferred a causal structure of two latent factors (Figure Xa), which we labelled *life history pace* (body size, female age at maturity, and longevity) and *niche complexity* (% fruit in diet, social group size, relative brain size).
+These factors are consistent with the observed fast-slow continuum of life history [@bielby2007fast; @healy2019animal], as well as the fact that frugivory is linked to sociality in primates [@wrangham1980ecological; @sterck1997evolution] and both are prominent predictors of relative brain size [@dunbar2017there; @decasien2017primate].
+We therefore analyzed the co-evolution of these two latent factors using our DPM.
+
+As shown in Figure Xb, we found that evolutionary changes in life-history pace (LH), shown along the x-axis, generate a co-evolutionary response in niche complexity (NC) (y-axis), as indicated by the diagonal blue line (the null-cline, where expected change in trait value = 0).
+Conversely, LH is unaffected by changes in NC, as indicated by the vertical black line, showing that LH stays near its average level (0) regardless of the value of NC.
+Thus, life-history pace is the leader, and niche complexity the follower in this co-evolutionary process.
+
+These results are consistent with the observations that repeated changes in size characterized mammalian evolution [@pagel2022general] and changes in body size associated with new niches were arguably primarily responsible for changes in relative brain size [@smaers2021evolution].
+Conversely, it appears moot to debate the relative importance of social and ecological factors in predicting relative brain size [@dunbar2017there; @decasien2017primate] as both are sides of the same encephelization coin.
+In short, as species evolve along the fast-slow continuum of life history (for exogenous reasons), selection acts to adjust their relative brain size, diet quality, and sociality in a coordinated fashion.
